
% !TeX encoding = UTF-8
% !TeX program = xelatex
% !TeX spellcheck = it_IT

\documentclass[noexaminfo,oneside,binding=0.6cm]{sapthesis}

\usepackage{enumitem}
\usepackage{polyglossia}
\usepackage{hyperref}
\usepackage[style=numeric]{biblatex}
\usepackage{csquotes}
\usepackage{subcaption}
\usepackage{amssymb}
\usepackage{float}
\usepackage{tabularx}
\usepackage{array}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{geometry}
\usepackage{linearb}
\usepackage{cypriot}
\usepackage{pifont}
\usepackage[linesnumbered,ruled,longend]{algorithm2e}
\usepackage{adjustbox}

\newcommand{\cmark}{\ding{51}} % ✓
\newcommand{\xmark}{\ding{55}} % ✗
\newcommand{\xxmark}{\ding{56}} % ✘ (heavier)

% --- Code listings ---
\usepackage{listings}        % main package for code
\usepackage{xcolor}          % colors for syntax highlighting
\usepackage[T1]{fontenc}     % proper encoding for monospace fonts
\usepackage{inconsolata}     % clean monospace font (better than default CM typewriter)

\setmainlanguage{english}
\setotherlanguage[variant=ancient]{greek}
\setotherlanguage{italian}
\newfontfamily\greekfont{Times New Roman}[Script=Greek]

\addbibresource{references.bib}


\hypersetup{pdftitle={An AI Framework for Linear B Translation into Ancient Greek and English},pdfauthor={Alessio Maiola}}


\title{An AI Framework for Linear B Translation into Ancient Greek and English}
\author{Alessio Maiola}
\IDnumber{1933744}
\course{Engineering in Computer Science}
\courseorganizer{Facoltà di Ingegneria dell'Informazione, Informatica e Statistica}
\AcademicYear{2024/2025}
\advisor{Prof. Aris Anagnostopoulos}
\authoremail{maiola.1933744@studenti.uniroma1.it}
\copyyear{2025}
\thesistype{Master Thesis}

\begin{document}

\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},   
    commentstyle=\color{codegreen},
    keywordstyle=\color{magenta},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\footnotesize\ttfamily,
    breakatwhitespace=false,         
    breaklines=true,                 
    captionpos=b,                    
    keepspaces=true,                 
    numbers=left,                    
    numbersep=5pt,                  
    showspaces=false,                
    showstringspaces=false,
    showtabs=false,                  
    tabsize=2
}

\lstset{style=mystyle}

\frontmatter
\maketitle
\dedication{Dedicated to \\ Luigi Ricci}


\begin{acknowledgments}
Voglio ringraziare tutte le persone che mi hanno supportato nel corso di questo progetto di tesi. \\
In particolare, il mio relatore Prof. Aris Anagnostopoulos per la sua guida e i suoi preziosi consigli. \\
Vorrei anche ringraziare i miei genitori, che mi hanno sempre incoraggiato a perseguire i miei interessi accademici. \\
Pur scettici di fronte alla complessità del tema, mi hanno sempre sostenuto e motivato a dare il meglio. \\
Vorrei ringraziare mio fratello, Emanuele, per ricordarmi di conoscere pur sempre il greco meglio di me, anche se non è vero. \\
Grazie ai miei nonni, che mi hanno sempre ispirato con la loro saggezza e il loro amore per la conoscenza. \\
Infine, un ringraziamento speciale va ai miei amici e colleghi. \\
Un grazie a Penelope, per aver contribuito al mio amore per il tema della tesi. \\
Grazie a Ludovica, Ilaria, Francesco e tutti gli altri per il supporto morale e le risate condivise durante questo percorso. \\
Grazie a QXIII per i momenti di svago e relax, che mi hanno permesso di ricaricare le energie. \\
Grazie a Marilena per avermi supportato e sopportato nei numerosi momenti di difficoltà incontrati. \\
Grazie a Gabriele, per avermi accompagnato in questo viaggio accademico e per il suo aiuto prezioso. \\
Grazie ai ragazzi di CyberChallenge.IT per avermi fatto scoprire un'altra passione, che mi ha aiutato a sviluppare capacità utili anche per questo progetto. \\
Infine un ringraziamento speciale al Prof. Tullio, per avermi introdotto al mondo della Grecia Antica e per avermi trasmesso la passione per questa materia.
\end{acknowledgments}


\begin{abstract}
This thesis tests whether large language models (LLMs) can assist in translating Linear B administrative texts into Ancient Greek and English.
I have built a compact \emph{Translation Pipeline} that couples brute-force cognate search with candidate aggregation, Linear-SVM auxiliary classifiers (part of speech, noun type, inflection), explicit handling of logograms/abbreviations, and structured multi-message prompting. Alongside the engineering, the thesis surveys the Aegean scripts (historical context, sites, linguistic features, and the decipherment of Linear B), details data collection (Linear A fragments, Linear B documents, Ancient Greek resources), and situates the approach within prior work on cognate matching and manual decipherment.

For cognates, I employed the generative \emph{NeuroDecipher} framework from Jiaming Luo and a brute-force extraction pass; for disambiguation, I have added auxiliary labels and table-driven guidance.
Using Gemini~2.5~Flash, the system reconstructs plausible Ancient Greek forms and produces English translations for full tablets.
Evaluation against translations from Tselentis' lexicon shows that automated translation is feasible: list-like inventories and administrative formulae are consistently recovered, vocabulary stays domain-appropriate, and English readings preserve the intended sense. An error taxonomy emerges: (i) weak/misleading cognates, (ii) case/number misassignment in context-poor lines, (iii) grammatical function errors from misclassification, and (iv) occasional logogram mishandling.
Greek reconstructions sometimes retain Mycenaean features, but these rarely obscure meaning.

I have released code and data and outline concrete next steps: hand-curating auxiliary labels, enriching the logogram lexicon, adding a normalization stage toward Classical Greek, fine-tuning on expert-validated pairs, broader evaluation, and cross-script generalization (e.g., Cypriot syllabary vs. alphabetic Greek).

\end{abstract}

\tableofcontents

\mainmatter

\include{introduction}
\include{history}
\include{cognates}
\include{classifiers}
\include{prompt}
\include{conclusion}

\backmatter
\cleardoublepage
\phantomsection % Give this command only if hyperref is loaded
\addcontentsline{toc}{chapter}{\bibname}
% Here put the code for the bibliography. You can use BibTeX or
% the BibLaTeX package or the simple environment thebibliography.

\printbibliography

\end{document}